XXX in a Tracing JIT Compiler for Efficient Dynamic Languages

### 2011-03-25 talk/icooolps2011/paper.tex
flesh out the introduction

It has long been an objective of the partial evaluation community to
automatically produce compilers from interpreters. There has been a recent
renaissance of this idea using the different technique of tracing just-in-time
compilers. A number of projects have attempted this approach. SPUR \cite{XXX}
is a tracing JIT for .NET together with a JavaScript implementation in C\#. XXX

All these projects have in common that they trace an implementation language
which is then used to implement an object model of a dynamic language. The
tracer then traces through this object model, which makes the object model
transparent to the tracer and its optimizations. Therefore the semantics of the
dynamic language does not have to be replicated in the JIT. We call this
approach \emph{meta-tracing}. We will give an introduction to the PyPy project
and to meta-tracing in Section~\ref{sec:Background}.

Another commonality of all these approaches is that they require some
annotations (or hints) in the dynamic language implementation to guide the
meta-tracer. SPUR and PyPy both provide the interpreter author with more
elaborate hints to influence the meta-tracer and its optimizer.

PyPy's hints go even further than SPUR's in that they provide the interpreter
author with a flexible toolset to make her implementation extremely efficient.
In this paper we present the two most prominent ones and show how classical
implementation techniques of dynamic languages can be expressed with them.

The contributions of this paper are:
\begin{itemize}
 \item A hint to introduce arbitrary constants into the trace.
 \item A way to define new pure operations which the optimizer then recognizes.
 \item A worked-out example of a simple object model of a dynamic language and how it can be improved using these hints.
\end{itemize}

@@

\section{Controlling Optimization}

@@

In this section we presented two more hints that can be used in the source code
of the interpreter. They are used to influence what the optimizer does with the
trace. The examples given here are a bit too small, the next
section gives a worked-out example that puts all the pieces together.

@@

\section{Putting It All Together}

@@

\section{Evaluation}
\label{sect:evaluation}

>>>>

It has long been an objective of the partial evaluation community to
automatically produce compilers from interpreters. There has been a recent
renaissance of this idea using the different technique of tracing just-in-time
compilers. A number of projects have attempted this approach. SPUR \cite{XXX} is
a tracing JIT for .NET together with a JavaScript implementation in C\#. PyPy
\cite{armin_rigo_pypys_2006} contains a tracing JIT for RPython (a restricted
subset of Python). This JIT is then used to trace a number of languages
implementations written in RPython. A number of other experiments in this
directions were done, such as an interpreter for Lua in JavaScript, which is run
on and optimized with a tracing JIT for JavaScript
\cite{yermolovich_optimization_2009}.

These projects have in common that they implement a dynamic language in some
implementation language. In addition they build a tracing JIT for that implementation
language. The tracing JIT then traces through the object model of the dynamic
language implementation. This makes the object model transparent to the tracer
and its optimizations. Therefore the semantics of the dynamic language does not
have to be replicated in a JIT. We call this approach \emph{meta-tracing}.
Another commonality of these approaches is that they allow some annotations (or
hints) in the dynamic language implementation to guide the meta-tracer. This
makes the process not completely automatic but can give good speedups over
bare meta-tracing.

In this paper we present two of these hints that are extensively used in the
PyPy project to improve the performance of its Python interpreter.
 The

PyPy's hints go even further than SPUR's in that they provide the interpreter
author with a flexible toolset to make her implementation extremely efficient.
In this paper we present the two most prominent ones and show how classical
implementation techniques of dynamic languages can be expressed with them. These
hints are used to control how the optimizer of the tracing JIT can improve the
traces of the object model. More specifically, these hints influence the
constant folding optimization. The first hint make it possible to turn arbitrary
variables in the trace into constants. The second hint allows the definition of
additional foldable operations.

Together these two hints can be used to express many classic implementation
techniques used for object models of dynamic languages, such as maps and
polymorphic inline caches.

The contributions of this paper are:
\begin{itemize}
 \item A hint to turn arbitrary variables into constants in the trace.
 \item A way to define new pure operations which the constant folding
 optimization then recognizes.
 \item A worked-out example of a simple object model of a dynamic language and
 how it can be improved using these hints.
\end{itemize}

The paper is structured as follows: Section~\ref{sec:background} gives an
introduction to the PyPy project and meta-tracing and presents an example of a
tiny dynamic language object model. Section~\ref{sec:hints} presents the hints,
what they do and how they are applied. Section~\ref{sec:fastobjmodel} shows how
the hints are applied to the tiny object model and Section~\ref{sec:evaluation}
presents benchmarks.

@@

\section{Hints for Controlling Optimization}
\label{sec:hints}

@@

In this section we presented two hints that can be used in the source code
of the interpreter. They are used to influence what the optimizer does with the
trace. The examples given here are a bit too small, the next
section gives a worked-out example that puts all the pieces together.

@@

\section{Putting It All Together}
\label{sec:fastobjmodel}

@@

\section{Evaluation}
\label{sec:evaluation}
